---
title: "Darwin Core mapping"
subtitle: "For: CamAlien"
author:
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This script prepares data for publication on [GBIF](https://www.gbif.org/). The script transforms the raw data to a [Darwin Core Archive](https://www.gbif.org/darwin-core) that can be harvested by GBIF. It was developed for the [Biodiversa+ pilot on invasive alien species monitoring](https://www.biodiversa.eu/2024/02/02/pilot-on-invasive-alien-species-monitoring-first-year-report/) project.

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load libraries

```{r}
library(here)
library(DBI)
library(duckdb)
library(dplyr)
library(readr)
```

## Load snapshot provided by Aarhus university

```{r}
file_location <- here::here("data", "raw", "detectionsummary_belgium_20250310_sample.csv")
snapshot <- readr::read_delim(file_location, show_col_types = FALSE, delim = ";")
```

## Set project-level values:

```{r}
license <- "CC0-1.0"
rights_holder <- "INBO"
dataset_name <- "Biodiversa+ IAS pilot - CamAlien observations of alien plant species in Flanders (Belgium)"
habitat <- "Riverine habitats"
camera_height_m <- 1.7
country_code = "BE"
```


# Create dwc

# Create occurrence

```{r}
occurrence <-
  snapshot %>% 
  dplyr::mutate(
    .keep = "none",
    type = "Image",
    license = license,
    rightsHolder = rights_holder,
    datasetID = "",
    collectionCode = "camALIEN",
    datasetName = dataset_name,
    basisOfRecord = "MachineObservation",
    occurrenceID = .data$detectionsummaryid,
    occurrenceStatus = "present",
    eventID = .data$imagedataid,
    parentEventID = "", #.data$deploymentID,
    eventDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"),
    habitat = habitat,
    samplingProtocol = "moving camera on vehicle",
    samplingEffort = "", #paste(
    #   format(.data$deploymentStart
    #, format = "%Y-%m-%dT%H:%M:%SZ"),
    #   format(.data$deploymentEnd, format = "%Y-%m-%dT%H:%M:%SZ"),
    #   sep = "/"
    # ),
    minimumDistanceAboveSurfaceInMeters = camera_height_m,
    maximumDistanceAboveSurfaceInMeters = camera_height_m,
    decimalLatitude = NA, #.data$latitude,
    decimalLongitude = NA, #.data$longitude,
    geodeticDatum = "EPSG:4326", # WGS84
    coordinateUncertaintyInMeters = NA, #TODO: fill in
    dateIdentified = format(
      .data$detection_timestamp,
      format = "%Y-%m-%dT%H:%M:%SZ"
    ),
    identificationRemarks = NA_character_,
    taxonID = paste0("https://www.gbif.org/species/", .data$gbifid),
    scientificName = .data$scientificname,
    kingdom = "Plantae"
  ) %>%
  dplyr::select(
    "type", "license", "rightsHolder", "datasetID", "collectionCode",
    "datasetName", "basisOfRecord", "occurrenceID", "occurrenceStatus",
    "eventID", "parentEventID", "eventDate", "habitat", "samplingProtocol", 
    "samplingEffort", "minimumDistanceAboveSurfaceInMeters",
    "maximumDistanceAboveSurfaceInMeters", "decimalLatitude", 
    "decimalLongitude", "geodeticDatum", "coordinateUncertaintyInMeters",
    "dateIdentified", "identificationRemarks", "taxonID", "scientificName", 
    "kingdom"
  )
```

# Create Audubon/Audiovisual Media Description extension

```{r}
multimedia <-
  snapshot %>%
  dplyr::arrange(.data$timestamp) %>%
  dplyr::mutate(
    .keep = "none",
    occurrenceID = .data$detectionsummaryid,
    identifier = .data$imagedataid,
    `dc:type` = "StillImage",
    `dcterms:rights` = "CC-BY-4.0",
    createDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"),
    captureDevice = NA_character_, # TODO: fill
    resourceCreationTechnique = "timelapse based on vehicle speed", # TODO: year-dependent
    accessURI = .data$url,
    `dc:format` = "image/jpeg", # TODO: year-dependent
    serviceExpectation = "online"
  ) %>%
  dplyr::select(
    "occurrenceID", "identifier", "dc:type", "dcterms:rights", "createDate",
    "captureDevice", "resourceCreationTechnique", "accessURI", "dc:format",
    "serviceExpectation",
  )
```
`
