---
title: "Darwin Core mapping"
subtitle: "For: camALIEN"
author:
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This script prepares data for publication on [GBIF](https://www.gbif.org/). The script transforms the raw data to a [Darwin Core Archive](https://www.gbif.org/darwin-core) that can be harvested by GBIF. It was developed for the [Biodiversa+ pilot on invasive alien species monitoring](https://www.biodiversa.eu/2024/02/02/pilot-on-invasive-alien-species-monitoring-first-year-report/) project.

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load libraries

```{r}
library(here)
library(DBI)
library(duckdb)
library(dplyr)
library(readr)
```
## Load DuckDB files

```{r}
file_location <- here::here("data", "raw", "camalien.duckdb")
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = file_location, read_only = TRUE)

detectionsummary <- DBI::dbGetQuery(con, "SELECT * FROM detectionsummary")
detection <- DBI::dbGetQuery(con, "SELECT * FROM detection") %>% 
  rename("detectionid" = "id")
plantnetcall <- DBI::dbGetQuery(con, "SELECT * FROM plantnetcall") %>% 
  dplyr::rename("callid" = "id")
species <- DBI::dbGetQuery(con, "SELECT * FROM species") %>% 
  dplyr::rename("speciesid" = "pnid")
image <- DBI::dbGetQuery(con, "SELECT * FROM image") %>% 
  dplyr::rename_with(
    ~ paste0("image", .x),
    c("id", "created", "filename", "path", "type")
    )
imagemeta <- DBI::dbGetQuery(con, "SELECT * FROM imagemeta") %>% 
  dplyr::rename("imagemetaid" = "id") %>% 
  dplyr::rename_with(~ paste0("image", .x), c("tag", "position", "timestamp"))
chunk <- DBI::dbGetQuery(con, "SELECT * FROM chunk") %>% 
  dplyr::rename("chunkid" = "id")
job <- DBI::dbGetQuery(con, "SELECT * FROM job") %>% # not needed
  dplyr::rename("jobid" = "id")
partner <- DBI::dbGetQuery(con, "SELECT * FROM partner") %>% 
  dplyr::rename(
    "partnerid" = "id",
    "partner" = "name")

# not needed?
plantnetcall_metadata <- 
  DBI::dbGetQuery(con, "SELECT * FROM plantnetcall_metadata")
plantnerrequestparams <- 
  DBI::dbGetQuery(con, "SELECT * FROM plantnetrequestparams") %>% 
  dplyr::rename("requestparamid" = "id")

DBI::dbDisconnect(con)
```

## Load snapshot provided by Aarhus university

```{r}
file_location <- here::here("data", "raw", "detectionsummary_belgium_20250310_sample.csv")
snapshot <- readr::read_csv(file_location, show_col_types = FALSE) %>% 
  mutate(
    imagetype = "tiff"
    )
```

# Create camalien dataset

```{r}
camalien <- #imagetype is extra
  dplyr::full_join(
    detection, detectionsummary, by = c("callid", "speciesid")
    ) %>% 
  dplyr::full_join(plantnetcall, by = "callid") %>% 
  dplyr::full_join(image, by = "imageid") %>% 
  dplyr::full_join(imagemeta, by = "imagemetaid") %>% 
  dplyr::full_join(chunk, by = "chunkid") %>% 
  dplyr::full_join(job, by = "jobid") %>% 
  dplyr::full_join(partner, by = "partnerid") %>% 
  dplyr::full_join(species, by = "speciesid") %>% 
  dplyr::select(
    partner, imageid, imagecreated, imagefilename, imagepath, imagetype, imagetag, 
    imageposition, imagetimestamp, velocity, altitude, callid, imageurl, 
    detectionid, maxscore, sumscore, count, coverage, gbifid, speciesid, 
    scientificname
    ) %>% 
  dplyr::rename("pnid" = "speciesid")
```

# Create dwc

# Create occurrence

```{r}
occurrence <-
  snapshot %>% 
  dplyr::mutate(
      .keep = "none",
      type = "Image",
      license = "CC0-1.0",
      rightsHolder = partner,
      datasetID = "",
      collectionCode = "camALIEN",
      datasetName = dataset_name,
      basisOfRecord = "MachineObservation",,
      occurrenceID = .data$detectionsummaryid,
      occurrenceStatus = "present",
      eventID = .data$imagedataid,
      #parentEventID = .data$deploymentID,
      eventDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"),
      habitat = "riverine habitats",
      samplingProtocol = "moving camera on vehicle",
      # samplingEffort = paste(
      #   format(.data$deploymentStart
      
      #, format = "%Y-%m-%dT%H:%M:%SZ"),
      #   format(.data$deploymentEnd, format = "%Y-%m-%dT%H:%M:%SZ"),
      #   sep = "/"
      # ),
      #eventRemarks = "x",
      minimumDistanceAboveSurfaceInMeters = camera_height,
      maximumDistanceAboveSurfaceInMeters = camera_height,
      decimalLatitude = .data$latitude,
      decimalLongitude = .data$longitude,
      geodeticDatum = "EPSG:4326", # WGS84
      #coordinateUncertaintyInMeters = .data$coordinateUncertainty,
      dateIdentified = format(
        .data$classificationTimestamp,
        format = "%Y-%m-%dT%H:%M:%SZ"
      ),
      identificationRemarks = paste0(
        # E.g. "classified by a machine with 89% certainty"
        dplyr::if_else(
          is.na(.data$classificationMethod),
          "classified",
          paste0("classified by a ", .data$classificationMethod)
        ),
        dplyr::if_else(
          is.na(.data$classificationProbability),
          "",
          paste0(" with ", .data$classificationProbability * 100, "% certainty")
        )
      ),
      taxonID = paste0("https://www.gbif.org/species/", .data$taxonID),
      scientificName = .data$scientificName,
      kingdom = "Plantae"
  )
    # ) %>%
    # dplyr::select(
    #   "type", "license", "rightsHolder", "datasetID", "collectionCode",
    #   "datasetName", "basisOfRecord", "dataGeneralizations", "occurrenceID",
    #   "individualCount", "sex", "lifeStage", "behavior", "occurrenceStatus",
    #   "occurrenceRemarks", "organismID", "eventID", "parentEventID",
    #   "eventDate", "habitat", "samplingProtocol", "samplingEffort",
    #   "eventRemarks", "locationID", "locality", "minimumDepthInMeters",
    #   "maximumDepthInMeters", "minimumDistanceAboveSurfaceInMeters",
    #   "maximumDistanceAboveSurfaceInMeters", "decimalLatitude",
    #   "decimalLongitude", "geodeticDatum", "coordinateUncertaintyInMeters",
    #   "coordinatePrecision", "identifiedBy", "dateIdentified",
    #   "identificationRemarks", "taxonID", "scientificName", "kingdom"
    )
```


```{r}
  # Create Audubon/Audiovisual Media Description extension
  multimedia <-
    snapshot %>%
    dplyr::arrange(.data$deploymentID, .data$timestamp, .data$fileName) %>%
    dplyr::mutate(
      .keep = "none",
      occurrenceID = .data$observationID,
      identifier = .data$mediaID,
      `dc:type` = dplyr::case_when(
        grepl("video", .data$fileMediatype) ~ "MovingImage",
        .default = "StillImage"
      ),
      comments = dplyr::case_when(
        !is.na(favorite) & !is.na(mediaComments) ~
          paste("marked as favorite", mediaComments, sep = " | "),
        !is.na(favorite) ~ "marked as favorite",
        .default = .data$mediaComments
      ),
      `dcterms:rights` = media_license,
      CreateDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"),
      captureDevice = .data$cameraModel,
      resourceCreationTechnique = dplyr::recode(
        .data$captureMethod,
        "activityDetection" = "activity detection",
        "timeLapse" = "time lapse"
      ),
      accessURI = .data$filePath,
      `dc:format` = .data$fileMediatype,
      serviceExpectation = dplyr::if_else(
        as.logical(.data$filePublic),
        "online",
        "authenticate"
      )
    ) %>%
    dplyr::select(
      "occurrenceID", "identifier", "dc:type", "comments", "dcterms:rights",
      "CreateDate", "captureDevice", "resourceCreationTechnique", "accessURI",
      "dc:format", "serviceExpectation",
    )
```


