---
title: "Darwin Core mapping"
subtitle: "For: CamAlien"
author:
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This script prepares data for publication on [GBIF](https://www.gbif.org/). The script transforms the raw data to a [Darwin Core Archive](https://www.gbif.org/darwin-core) that can be harvested by GBIF. It was developed for the [Biodiversa+ pilot on invasive alien species monitoring](https://www.biodiversa.eu/2024/02/02/pilot-on-invasive-alien-species-monitoring-first-year-report/) project.

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load libraries

```{r load libraries}
library(here)
library(DBI)
library(duckdb)
library(dplyr)
library(readr)
library(tidyverse)
```

## Load snapshot provided by Aarhus university

```{r import data}
file_location <- here::here("data", "raw", "detectionsummary_belgium_20250310_sample.csv")
snapshot <- readr::read_delim(file_location, show_col_types = FALSE, delim = ";")
```

## Import project-level data

```{r import google sheets}
project_raw <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQzj3LUe1Tbd3UNf9OiFQeKBNKmNh4nB3Tb0m6tQVRWxE9r9CQZ0bZuVAdmaOIU4ciYQzeupbUA2oR7/pub?gid=1501598855&single=true&output=csv", show_col_types = FALSE) 
metadata_raw <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQzj3LUe1Tbd3UNf9OiFQeKBNKmNh4nB3Tb0m6tQVRWxE9r9CQZ0bZuVAdmaOIU4ciYQzeupbUA2oR7/pub?gid=0&single=true&output=csv", show_col_types = FALSE)
creators_raw <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQzj3LUe1Tbd3UNf9OiFQeKBNKmNh4nB3Tb0m6tQVRWxE9r9CQZ0bZuVAdmaOIU4ciYQzeupbUA2oR7/pub?gid=1358621000&single=true&output=csv", show_col_types = FALSE)
```

## Prepare project-level (meta)data

```{r}
project <-
  project_raw %>% 
  dplyr::select(term, value) %>% 
  tidyr::pivot_wider(names_from = term, values_from = value)
metadata <- 
  metadata_raw %>% 
  dplyr::select(term, value) %>% 
  tidyr::pivot_wider(names_from = term, values_from = value)
creators <-
  creators_raw %>% 
  dplyr::arrange(order) %>% 
  dplyr::mutate(title = paste(first_name, last_name))
creators_list <- purrr::map(1:nrow(creators), function(i) {
  list(
    title = creators$title[i],
    first_name = creators$first_name[i],
    last_name = creators$last_name[i],
    organization = creators$organization[i],
    country = creators$country[i],
    email = creators$email[i],
    orcid = creators$orcid[i]
  )
})

metadata_provider_names <- 
  creators %>% 
  dplyr::filter(metadata_provider == TRUE) %>%
  dplyr::pull(title)
metadata_provider <- 
  purrr::keep(creators_list, ~ .x$title %in% metadata_provider_names)

contact_names <- 
  creators %>% 
  dplyr::filter(contact == TRUE) %>%
  dplyr::pull(title)
contact <- purrr::keep(creators_list, ~ .x$title %in% contact_names)

project_personnel_names <- 
  creators %>% 
  dplyr::filter(project_personnel == TRUE) %>% 
  dplyr::pull(title)
project_personnel <- 
  purrr::keep(creators_list, ~ .x$title %in% project_personnel_names)
```

## Prepare data

```{r}
snapshot <-
  snapshot %>% 
  dplyr::mutate(
    timestamp_date = as.Date(.data$timestamp, tz = "UTC"),
    year = format(.data$timestamp_date, format = "%Y"),
    deploymentID = paste(metadata$country_code, timestamp_date, sep = "_")
  )
deployments <-
  snapshot %>% 
  dplyr::group_by(deploymentID) %>% 
  summarise(
    deployment_start = min(.data$timestamp),
    deployment_end = max(.data$timestamp)
  )
snapshot <-
  snapshot %>%
  dplyr::left_join(deployments)
```


# Create dwc

## Create occurrence

```{r occurrence}
occurrence <-
  snapshot %>% 
  dplyr::mutate(
    .keep = "none",
    type = "Image",
    license = metadata$license_data,
    rightsHolder = metadata$rights_holder,
    datasetID = "",
    collectionCode = "CamAlien",
    datasetName = metadata$title,
    basisOfRecord = "MachineObservation",
    occurrenceID = .data$detectionsummaryid,
    occurrenceStatus = "present",
    eventID = .data$imagedataid,
    parentEventID = .data$deploymentID,
    eventDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"), #TODO: check time zones
    habitat = metadata$habitat,
    samplingProtocol = "moving camera on vehicle",
    samplingEffort = paste(
      format(.data$deployment_start, format = "%Y-%m-%dT%H:%M:%SZ"),
      format(.data$deployment_end, format = "%Y-%m-%dT%H:%M:%SZ"),
      sep = "/"
    ),
    minimumDistanceAboveSurfaceInMeters = metadata$camera_height_m,
    maximumDistanceAboveSurfaceInMeters = metadata$camera_height_m,
    decimalLatitude = NA, #.data$latitude,
    decimalLongitude = NA, #.data$longitude,
    geodeticDatum = "EPSG:4326", # WGS84
    coordinateUncertaintyInMeters = NA, #TODO: fill in
    dateIdentified = format(
      .data$detection_timestamp,
      format = "%Y-%m-%dT%H:%M:%SZ"
    ),
    identificationRemarks = NA_character_, #TODO: fill in
    taxonID = paste0("https://www.gbif.org/species/", .data$gbifid),
    scientificName = .data$scientificname,
    kingdom = "Plantae"
  ) %>%
  dplyr::select(
    "type", "license", "rightsHolder", "datasetID", "collectionCode",
    "datasetName", "basisOfRecord", "occurrenceID", "occurrenceStatus",
    "eventID", "parentEventID", "eventDate", "habitat", "samplingProtocol", 
    "samplingEffort", "minimumDistanceAboveSurfaceInMeters",
    "maximumDistanceAboveSurfaceInMeters", "decimalLatitude", 
    "decimalLongitude", "geodeticDatum", "coordinateUncertaintyInMeters",
    "dateIdentified", "identificationRemarks", "taxonID", "scientificName", 
    "kingdom"
  )
```

## Create Audubon/Audiovisual Media Description extension

```{r multimedia}
multimedia <-
  snapshot %>%
  dplyr::arrange(.data$timestamp) %>%
  dplyr::mutate(
    .keep = "none",
    occurrenceID = .data$detectionsummaryid,
    identifier = .data$imagedataid,
    `dc:type` = "StillImage",
    `dcterms:rights` = "CC-BY-4.0",
    createDate = format(.data$timestamp, format = "%Y-%m-%dT%H:%M:%SZ"),
    captureDevice = NA_character_, # TODO: fill in
    resourceCreationTechnique = dplyr::if_else(
      .data$year == 2024,
      "fixed sampling rate",
      "timelapse based on vehicle speed"
      ),
    accessURI = .data$url,
    `dc:format` = "image/jpeg", # TODO: year-dependent
    serviceExpectation = "online"
  ) %>%
  dplyr::select(
    "occurrenceID", "identifier", "dc:type", "dcterms:rights", "createDate",
    "captureDevice", "resourceCreationTechnique", "accessURI", "dc:format",
    "serviceExpectation",
  )
```

# Create metadata file (eml.xml)

## Function to create contributors in EML format

```{r}
create_eml_contributors <- function(contributor_list) {
  purrr::map(contributor_list, ~ EML::set_responsibleParty(
    givenName = .$first_name,
    surName = .$last_name,
    organizationName = .$organization, # Discouraged by EML, but used by IPT
    email = .$email,
    userId = if (!is.na(.$orcid)) {
      list(directory = "https://orcid.org/", .$orcid)
    } else {
      NULL
    }
  ))
}
```

This script is based on input data above.

```{r EML initiation}
# Initiate EML
eml <- list(
  packageId = uuid::UUIDgenerate(),
  system = "uuid",
  dataset = list()
)
```

```{r EML project}
eml$dataset$project$title <- project$title
eml$dataset$project$abstract$para <- project$description
eml$dataset$project$funding$para <- project$funding
eml$dataset$project$studyAreaDescription$descriptor$descriptorValue <-
  project$study_area_description
eml$dataset$project$designDescription$description$para <- 
  project$design_description
eml$dataset$project$personnel <- create_eml_contributors(project_personnel)
```

```{r EML dataset}
# Set short name and title
eml$dataset$shortName <- metadata$short_name
eml$dataset$title <- metadata$title

# Set abstract
eml$dataset$abstract$para <- 
  c(
    metadata$description_paragraph1,
    metadata$description_paragraph2,
    metadata$description_paragraph3,
    metadata$description_paragraph4
    )

# Set update frequency (requires a description, even if empty)
eml$dataset$maintenance <- list(
  description = list(para = ""),
  maintenanceUpdateFrequency = metadata$update_frequency
)

# Set creators
eml$dataset$creator <- create_eml_contributors(creators_list)

# Set contacts
eml$dataset$contact <- create_eml_contributors(contact)

# Set metadata providers
eml$dataset$metadataProvider <- create_eml_contributors(metadata_provider)

# Set keywords
eml$dataset$keywordSet <-
  list(
    list(
      keywordThesaurus = paste(
        "GBIF Dataset Type Vocabulary:",
        "http://rs.gbif.org/vocabulary/gbif/dataset_type_2015-07-10.xml"
      ),
      keyword = "Occurrence"
    ),
    list(
        keywordThesaurus = paste(
          "GBIF Dataset Subtype Vocabulary:",
          "http://rs.gbif.org/vocabulary/gbif/dataset_subtype.xml"
        ),
        keyword = "Observation"
      ),
    list(
      keywordThesaurus = "n/a",
      keyword = metadata$keywords
    )
  )

# Set license
eml$dataset$intellectualRights$para <- metadata$license_data

# Set temporal and geographic coverage and geographicDescription
eml$dataset$coverage <-
  EML::set_coverage(
    beginDate = format(min(as.Date(occurrence$eventDate)), format = "%Y-%m-%d"),
    endDate = format(max(as.Date(occurrence$eventDate)), format = "%Y-%m-%d"),
    geographicDescription = metadata$geographic_description,
    westBoundingCoordinate = min(occurrence$decimalLongitude),
    southBoundingCoordinate = min(occurrence$decimalLatitude),
    eastBoundingCoordinate = max(occurrence$decimalLongitude),
    northBoundingCoordinate = max(occurrence$decimalLatitude)
  )

# Set taxonomic coverage
taxonomic_coverage <- data.frame(
  scientificName = "Plantae",
  taxonRank = "kingdom"
)
eml$dataset$coverage$taxonomicCoverage <-
  list(
    taxonomicClassification =
      purrr::map(1:nrow(taxonomic_coverage), function(i) {
        current_row <- taxonomic_coverage[i, ]
        list(
          taxonRankName = current_row$taxonRank,
          taxonRankValue = current_row$scientificName
        )
      })
  )
eml$dataset$coverage$taxonomicCoverage$generalTaxonomicCoverage <- metadata$taxonomic_description
```

```{r EML additional metadata}
# Bibliographic citations
#eml$additionalMetadata$metadata$gbif$bibliography$citation <- references
```

# Write files

```{r write files}
dir_export <- here::here("data", "processed", metadata$country_code)
if (!dir.exists(dir_export)) {
    dir.create(dir_export, recursive = TRUE)
  }
readr::write_csv(occurrence, file.path(dir_export, "occurrence.csv"), na = "")
readr::write_csv(multimedia, file.path(dir_export, "multimedia.csv"), na = "")
EML::write_eml(eml, file.path(dir_export, "eml.xml"))
```

# Validate EML

```{r}
EML::eml_validate(file.path(dir_export, "eml.xml"))
```

